---
title: "gtrendsR tutorial for beginners - Manchester R User group"
author: "Charlotte Woolley"
date: "25th of March 2024"
output: 
  html_document:
    code_folding: show
---

The purpose of this tutorial is to give a brief introduction to Google Trends data
using the gtrendsR package and to demonstrate its potential application within the field of biology. 

## **Introduction to Google Trends**

### What is Google Trends?
Google Trends is a freely available and publically accessible online facility of Google Inc. that provides pre-processed search query data from January the 1st 2004 until present day. The Google Trends [website](https://trends.google.co.uk/trends/) provides data and regularly updated trending themes/topics.

### Why is internet search query data important?

In recent times, the internet has become an essential source of information in many parts of the world. Over 50% of the world's population now has internet access and in Western countries such as the UK, internet usage has reached 88% of the population (Office for National Statistics, 2016).
Clearly, the internet isn't getting any less important and will no doubt be relied on increasingly
for a plethora of information. Indeed, increased reliance on the internet has led to the emergence of a type of research which falls under the term ‘infodemiology” coined by Eysenbach (2002), which essentially involves the study of the internet to reveal the distribution and determinants of health information, with the aim of improving public health.

### What can internet search query data be used for? 

Many professionals, such as scientists, journalists, marketers, business and finance specialists, politicians and health care workers have realised the potential of using internet search queries as a source of useful information. Google Trends has been used for a wide range of scientific research, including:

* General health, e.g Multiple sclerosis (Moccia *et al*., 2016)
* Infectious disease, e.g Zika (Dimitrov *et al*., 2016)
* Mental health, e.g Depression (Yang *et al*., 2010; 2011)
* Veterinary health, e.g Vomiting outbreak in dogs (Woolley *et al*., 2021)
* Financial, e.g Quantifying trading behaviour (Preis *et al*., 2013)
* Marketing, e.g Car sales (Carrière-Swallow & Labbé, 2013)
* Political, eg. Predicting elections (Granka, 2013)

### What exactly is Google Trends data?

Google Trends data is not a huge dataset containing every search made by every IP address in the world. To ensure that it is practical and ethical to use the information obtained from search queries, Google pre-processes the data before it is released, so that it is:

* Normalised; it is an ‘index’ of the proportion of queries within a particular geographical region at a particular time divided by the total number of queries at that region and time, where the data is scaled between 0 to 100 (to prevent places with the most search volume would always being ranked the highest).
* Anonymised; an unreported privacy threshold is kept so that if the total number of searches for a particular time and region are less than this value, a zero will be reported
* Aggregated; for a time series of up to 4 hours - data is supplied for past 4 hours - every minute, 24 hours - every 8 mins, 7 days - every hour, 90 days - every day, 5 years - every week, more than 5 years - every month
* Cached daily; the same request on the same date will return the same data
* Rounded to the nearest integer
* Categorised into about 25 main categories and hundreds of subcategories
* De-duplicated so that the same person making several of the same requests over a short period of time will not influence the results
* De-symbolised so that all special characters are filtered out of the search term

(Google, 2024; Milinovich *et al*., 2014; Stephens-Davidowitz & Varian, 2015) 

### Related terms and topics
In addition to providing temporal and locational data, Google also provides information of the top and rising 'related queries' and 'related topics' 

* Related queries (sometimes referred to as terms) are matches within your query in the same language. For example, if your search term is 'coconut', results will include 'coconut milk'. If your search term is 'coconut milk' results will include 'coconut oil' and 'pasteurised milk' 
* Related topics are a group of terms that have a similar concept and do not have to be in the same language. For example, if your search term is 'Scotland', results will include 'Schottland' (German for Scotland) and 'United Kingdom' (related topic) 
* Top refers to the most frequent and rising refers to the most growth in volume. 

(Google, 2024) 

### Understanding potential biases in Google Trends data

Despite the fact it is an undeniably useful tool, Google Trends data is not an ultimately superior source of information and contains its own individual biases that must be considered, like all data. Some important things to consider include:

* The effect of the media; Google Trends data was positively correlated with the number of Ebola cases during the outbreak, but it was prone to distortion due to lack of digital access in developing countries and differences in media coverage (Alicino *et al*., 2015)
* Unanticipated changes in search algorithm dynamics over time; shifts in linguistics, culture, technological changes and other unforeseen factors can influence search algorithms
* The population of internet users; the internet was initially more available to those in academic institutions and/or populated areas and now has much broader use by the wider population
* Search query selection; there are many possible terms that could be selected and ambiguity of search terms can cause interpretation errors if there is more than one meaning or spelling for a word 
* Don't be deceived by big data hubris (the assumption that big data substitute rather than compliment traditional data sources); the more sources of data you have, the more reliable your models will be!
* Lack of data supplied by Google Trends; sometimes there will not be enough data from a particular region to bypass Google's anonymity threshold, which can cause certain regions to appear disproportionately uninterested in a particular search query - low volume searches appear as 0
* Interference from Google suggestions; when people are searching, they might change their natural searching behaviour based on Google search suggestions or spelling corrections
* Changes in the Google data collection system; Google improved their data collection systems on the 01/01/2022, 01/01/2017 and 01/01/2016 and improved their geographical assignment on the 01/01/2011
* Language differences in keywords can make it difficult to decide which ones to use - if you use an English word this does not necessarily represent people who speak different languages
* A recent paper by Franzen (2023) entitled "Big data, big problems: Why scientists should refrain from using Google Trends" demonstrates a clear lack of understanding of the biases underlying Google Trends and when the data is likely to be, and not to be, reliable.

(Nuti *et al*., 2014; Stephens-Davidowitz & Varian, 2015; Google, 2024)


## **Installing and understanding the gtrendsR package**

First if all, install the gtrendsR package (Massicotte & Eddelbuettel, 2016).
```{r, eval=FALSE}
install.packages("gtrendsR")
```

Load gtrendsR and any other useful packages for this tutorial. Set options for scientific notation.
```{r, message=FALSE}
library(gtrendsR)
library(tidyverse)
library(ggmap)
library(lubridate)

options(scipen = 999)
```

The gtrendsR package is mainly built around the 'gtrends' function. This function contains the same arguments as can be specified for retrieving data on the Google Trends website. It is important to understand what these mean so I recommend checking out the help file (using ?gtrends). In order to know which codes to enter into the gtrends function for location and category, you need to extract these out of the package and view them. Locations are divided into countries and regions and categories are similarly divided into major catgories and subcategories. 
```{r}
data("countries")
head(countries)
head(subset(countries, countries$country_code =='GB'), 15)
data("categories")
head(categories)
```

## **Using Google Trends to explore temporal and geographical patterns in search query data**

### Example 1 - Investigate worldwide searches for 'menopause' over the past week

Use the gtrends function with the appropriate arguements. 
```{r}
menopause <- gtrends(keyword = 'menopause', time = 'now 7-d')
```

The function produces an object that is a list of 7 dataframes. Dataframes are 
'none' if Google Trends cannot produce enough data to bypass the Google 
confidentiality rules or if they are not applicable.
```{r}
summary(menopause)  
```

To take an initial look at the data over the course of the week, the data does
not need to be extracted from the list. 
```{r}
plot(menopause)
```

However, this is not the 'prettiest' graph and it would be interesting to view the
results summarised by day. To do this, the 'interest over time' dataframe needs to be extracted
from the gtrends object.
```{r}
menopause_iot <- as.data.frame(menopause$interest_over_time)

menopause_iot <- menopause_iot %>%
mutate(date = as.Date(date),
        day = factor(lubridate::wday(ymd(date), label = TRUE),
        levels = c('Mon', 'Tue', 'Wed', 'Thu', 'Fri', 'Sat', 'Sun')))

ggplot(menopause_iot, aes(x = day, y = hits, group = day)) +
        geom_boxplot() +
        stat_summary(fun.y=mean, colour="red", geom="point", shape=18, size=3.5) +
        labs(title = "Google searches for 'menopause' over the past week", 
             x = "Day", 
             y = "Search ratio") +
theme_bw()
```


### Example 2 - Investigate UK searches for 'menopause' and 'HRT' over the past 5 years

Use the gtrends function with the appropriate arguements. 
```{r}
menopause2 <- gtrends(keyword = c('menopause', 'HRT'), geo = 'GB', time = 'today+5-y')
```

Now we can extract the 'interest over time' database and plot it using ggplot
```{r}
menopause2_iot <- as.data.frame(menopause2$interest_over_time)
menopause2_iot$date <- as.Date(menopause2_iot$date)
ggplot(menopause2_iot, aes(x=date, y = hits,  colour = keyword)) +
        geom_line() +
        labs(title = "Google searches in Great Britain for 'menopause' in the last 5 years", 
             x = "Year", 
             y = "Search index",
             colour = "Keyword") +
        scale_y_continuous(breaks=c(seq(from=0, to=100, by=10))) +
        scale_x_date(date_breaks="1 year", date_labels = "%Y") +
        theme_bw() +
        theme(axis.text.x=element_text(angle=90,hjust=1))
```

The Davina McAll effect
BBC Breakfast "Wake up to menopause" - Monday 13th May- Friday 17th May 2019
"Sex Myths and Menopause 1" - 12th May 2021
"Sex Myths and Menopause 2" - 2nd May 2022
World menopause day - 18 October 2022


It would be interesting to look at which region within the UK searches for
'menopause' the most. To do this, the 'interest by region' dataframe needs to be extracted.
```{r}
menopause2_ibr <- as.data.frame(menopause2$interest_by_region) %>%
  filter(keyword == "menopause")
```

Then we can sort the levels of the regions in order of the largest search ratio
and make a bar graph of the data using ggplot.
```{r}
menopause2_ibr$location <- factor(menopause2_ibr$location, 
levels = menopause2_ibr$location[order(menopause2_ibr$hits, decreasing = TRUE)])

ggplot(menopause2_ibr, aes(x=location, y = hits)) +
        geom_bar(stat='identity', fill = c('purple4','forestgreen', 'blue4', 'red4')) +
        labs(title = "Google searches in Great Britain for 'menopause' over the last 5 years", 
             x = "Region", 
             y = "Search index") +
        scale_y_continuous(breaks=c(seq(from=0, to=100, by=10))) +
theme_bw()
```

We can also have a look at the related queries to 'menopause'. In this
tutorial I am going to focus on the top (or most frequent) related queries
but it is important to note that it is also possible to view the rising (or most rapidly
increasing) related queries and information about related topics too.
```{r}
menopause2_rq <- as.data.frame(menopause2$related_queries) %>%
  filter(keyword == "menopause")

top_queries <- subset(menopause2_rq, menopause2_rq$related_queries == 'top')
top_queries$subject <- as.numeric(top_queries$subject)
top_queries$value <- factor(top_queries$value, 
                        levels = top_queries$value[order(top_queries$subject, decreasing = TRUE)])
ggplot(top_queries, aes(x=value, y = subject)) +
        geom_bar(stat='identity') +
        labs(title = "Top related Google queries to 'menopause' in the UK over the last 5 years", 
             x = "Related queries", 
             y = "Search index") +
        scale_y_continuous(breaks=c(seq(from=0, to=100, by=10))) +
        theme_bw() +
        theme(axis.text.x=element_text(angle=90,hjust=1))
```


### Example 3 - Investigate worldwide searches for 'menopause' since 2010

Use the gtrends function with the appropriate arguements. 
```{r}
menopause3 <- gtrends(keyword = 'menopause', time = "2010-01-01 2024-01-01")
```

Now we can extract the 'interest over time' database and plot it using ggplot
```{r}
menopause3_iot <- as.data.frame(menopause3$interest_over_time)
menopause3_iot$date <- as.Date(menopause3_iot$date)
menopause3_iot$time <- 1:nrow(menopause3_iot)
ggplot(menopause3_iot, aes(x=date, y = hits)) +
        geom_line() +
        labs(title = "Google searches worldwide for 'menopause' since 2010", 
             x = "Year", 
             y = "Search index") +
        scale_y_continuous(breaks=c(seq(from=0, to=100, by=10))) +
        scale_x_date(date_breaks="1 year", date_labels = "%Y") +
        theme_bw() +
        theme(axis.text.x=element_text(angle=90,hjust=1))
```


There is an obvious increase in interest in 'menopause'! Now to see which countries are most interested...
```{r}
menopause3_ibr <- as.data.frame(menopause3$interest_by_country)
menopause3_ibr$location <- factor(menopause3_ibr$location, 
                                levels = menopause3_ibr$location[order(as.numeric(menopause3_ibr$hits), decreasing = TRUE)])

ggplot(menopause3_ibr %>% filter(hits != ""), aes(x=location, y = as.numeric(hits))) +
        geom_bar(stat='identity') +
        labs(title = "Google searches for 'menopause' worldwide", 
             x = "Region", 
             y = "Search index") +
        scale_y_continuous(breaks=c(seq(from=0, to=100, by=10))) +
        theme_bw() +
        theme(axis.text.x=element_text(angle=90,hjust=1))
```
Now, to view this data in a heat map of the world it is quite easy because the
gtrendsR package already provides country names. However, a few of the country names between
map data and gtrends data need changing for consistency (see below). 
```{r}
menopause3_ibr$region <- ifelse(menopause3_ibr$location == 'United States','USA', 
                                ifelse(menopause3_ibr$location == 'United Kingdom', 'UK', 
                                        ifelse(menopause3_ibr$location == 'Czechia', 'Czech Republic', 
                                                as.character(menopause3_ibr$location))))
```

Now we just need to extract worldwide map data from ggmap, categorise the 
search hits (I have used categories of 20) and plot the data using ggplot
```{r, message=FALSE}
new_map <- full_join(map_data("world"), menopause3_ibr)
new_map$search_index <- factor(ifelse(is.na(new_map$hits), 'no data', 
                                ifelse(new_map$hits < 20, '< 20', 
                                 ifelse(new_map$hits >= 20 & new_map$hits < 40, '20 - 39', 
                                  ifelse(new_map$hits >= 40 & new_map$hits < 60, '40 - 59', 
                                   ifelse(new_map$hits >= 60 & new_map$hits < 80, '60 - 79', 
                                    ifelse(new_map$hits >= 80, '> 80', new_map$hits)))))), 
                                        levels = c('> 80', '60 - 79', '40 - 59', '20 - 39', '< 20', 'no data'))
ggplot(new_map, aes(x=long, y=lat, group=group, fill=search_index)) + 
        geom_polygon(colour="black", show.legend = T) +
        scale_fill_manual(values= alpha(c("darkred", "orangered3", "orange2", 
                                          "yellow1", "khaki1", "grey50"), 0.5)) +
        labs(title = "Google searches for 'menopause' worldwide", 
             x = "Longitude", 
             y = "Latitude") +
        theme_bw()
```


## **Final remarks**

This tutorial was intended to demonstrate some of the potential uses and applications
of the gtrendsR package and give you the tools to perform basic data visualisations. 
There are many other applications of gtrendsR and this tutorial barely scratched
the surface.

You can find the full tutorial on my github:
https://github.com/CharlotteWoolley/Google_Trends_gtrendsR_Tutorial

### Get Trending!


## **References**

* Carrière-Swallow, Y., Labbé, F. (2011). Nowcasting with Google Trends in an Emerging Market. Journal of forecasting, 32, 289-298.
* Dimitrov, B. D. (2016). Nowcasting incidence of emergent zika virus infection and its outbreaks using Google Trends data: Examples from Brazil and Colombia. International Journal of Infectious Diseases, 53, 98.
* Eysenbach, G. (2002). Infodemiology: the epidemiology of (mis)information. The American Journal of Medicine, 113(9), 763-765
* Franzén, A. (2023). Big data, big problems: Why scientists should refrain from using Google Trends. Acta Sociologica, 66(3), 343-347. 
* Google. (2024). Google Trends Help. Retrieved January 10, 2017, from https://support.google.com/trends/?hl=en#topic=6248052
* Granka, L. (2013). Using Online Search Traffic to Predict US Presidential Elections. PS: Political Science & Politics, 46(2), 271-279
* Massicotte, P., Eddelbuettel, D. (2016). Package ‘gtrendsR’. Retrieved 30 March, 2017, from https://cran.rproject.org/web/packages/gtrendsR/gtrendsR.pdf
* Milinovich, G. J., Avril, S. M., Clements, A. C., Brownstein, J. S., Tong, S., & Hu, W. (2014). Using internet search queries for infectious disease surveillance: screening diseases for suitability. BMC Infectious Diseases, 14, 690
* Moccia, M., Palladino, R., Falco, A., Saccà, F., Lanzillo, R., & Brescia Morra, V. (2016). Google Trends: new evidence for seasonality of multiple sclerosis. Journal of Neurology, Neurosurgery, and Psychiatry, (1), jnnp-2016-313260. 
* Nuti, S. V., Wayda, B., Ranasinghe, I., Wang, S., Dreyer, R. P., Chen, S. I., & Murugiah, K. (2014). The use of google trends in health care research: A systematic review. PLoS ONE, 9(10). 
* Office for National Statistics. (2016). Internet users in the UK: 2016. Statistical Bulletin, 1–9. Retrieved 08 January, 2017, from https://www.ons.gov.uk/businessindustryandtrade/itandinternetindustry/bulletins/internetusers/2016
* Preis, T., Moat, H. S., Stanley. H. E. (2013). Quantifying trading behavior in financial markets using Google Trends. Scientific Reports, 3, 1684
* Stephens-Davidowitz, S., & Varian, H. (2015). A Hands-on Guide to Google Data. Google, Inc., 1–25. Retrieved January 10, 2017, from http://people.ischool.berkeley.edu/~hal/Papers/2015/primer.pdf
* Woolley, C. S. C., Handel, I. G., Bronsvoort, B. M., Schoenebeck, J. J., Clements, D. N. Surveillance of a vomiting outbreak in dogs in the UK using owner-derived and internet search data. (2021). Vet Rec. e308. https://doi.org/10.1002/vetr.308
* Yang, A. C., Huang, N. E., Peng, C. K., & Tsai, S. J. (2010). Do seasons have an influence on the incidence of depression? the use of an internet search engine query data as a proxy of human affect. PLoS ONE, 5(10), 1–7. 
* Yang, A. C., Tsai, S. J., Huang, N. E., & Peng, C. K. (2011). Association of Internet search trends with suicide death in Taipei City, Taiwan, 2004-2009. Journal of Affective Disorders, 132(1–2), 179–184. 

